Staged Abstract Interpreters

### 2019-08-15 paper/80_related.tex
typos

Splitting an analysis into multiple stages is also studied for
analysis other than abstract-interpretation-based control-flow
analysis, though the formulation may very
different. \citet{DBLP:conf/cgo/HardekopfL11} apply staging to a
flow-sensitive pointer analysis. The first stage is to analyze the
program code to obtain a sparse representation, and then the second
stage conducts the flow-sensitive analysis by refining the first
one. Another large body of static analyses are implemented by
generating constraints firstly and then let a solver discharge these
proof obligations or find the fixed-point. Notable examples include
SAT- and SMT-based program analysis
\cite{Gulwani:2008:PAC:1375581.1375616} and Datalog-based points-to
analysis \cite{Smaragdakis:2015:PA:2802194.2802195}.
Similar to the idea in this paper, such analyses can be splitted to
staged.  One of the future work is to explore that whether these
analyses can be implemented and optimized from
metaprogramming. Souffl{\'e}, one of the state-of-the-art Datalog
engine that is commonly used in points-to analysis
\cite{Antoniadis:2017:PDS:3088515.3088522}, uses the idea of Futamura
projection for efficent implemetation \cite{10.1007/978-3-319-41540-6_23}.
But it only specializes Datalog rules to the running engine, instead
of a full pipeline from program AST to generating constraints and
solving constraints.

Recent work shows that the numerical abstract domains can be
specialized w.r.t. the static program structure, although not in the
way of using staging. For example, decomposing polyhedras
\cite{DBLP:conf/popl/SinghPV17, Singh:2017:PCD:3177123.3158143} to
smaller ones that depend on the variables involved in a statement can
significantly improve the running time and space of abstract
transformers.

@@

\paragraph{Control-flow Analysis} Our case studies and evaluation focus on
control-flow anslysis for functional languages
\cite{Shivers:1991:SSC:115865.115884, Midtgaard:2012:CAF:2187671.2187672}. In
general, $k$-CFA (where $k > 0$) is considered impractical to use as a compiler
pass, due to its EXPTIME lower bound \cite{VanHorn:2008:DKC:1411204.1411243}.
However, several variants of 0CFA were developed and shown to be useful in
practice and program optimization \cite{Adams:2011:FTR:2048066.2048105,
Bergstrom:2014:PEH:2628136.2628153, ashley:practical, Reppy:2006:TCA:1159876.1159888}.

\paragraph{Two-level Semantics} The idea of reinterpreting the semantics as
abstract interpretation can be traced to Nielson's two-level semantics
\cite{NIELSON1989117}; using two-level semantics for code generation
was also explored by \citet{NIELSON198859}.
\citet{Sergey:2013:MAI:2491956.2491979}'s work of monadic abstract
interpreters is also closely related to the two-level semantics: the
use of a generic interface with monads and then reinterpret it by
different semantics is already two-level. Instead of focusing on
semantics, this paper shows how a staged analyzer can be built and
used to increase efficieny of static analysis, we augment the monadic
abstract interpreter from another dimension, using code generation to
produce efficient low-level code. Hence the presented work can be
considered as an two-dimension application of the two-level semantics.

>>>>

Splitting an analysis into multiple stages is also studied for
analysis other than abstract-interpretation-based control-flow
analysis, though the formulation may very
different. \citet{DBLP:conf/cgo/HardekopfL11} apply staging to a
flow-sensitive pointer analysis. The first stage is to analyze the
program code to obtain a sparse representation, and then the second
stage conducts the flow-sensitive analysis by refining the first
one. Another large body of static analyses are implemented by
generating constraints firstly and then let a solver discharge these
proof obligations or find the fixed-point. Notable examples include
SAT- and SMT-based program analysis
\cite{Gulwani:2008:PAC:1375581.1375616} and Datalog-based points-to
analysis \cite{Smaragdakis:2015:PA:2802194.2802195}.
Similar to the idea in this paper, such analyses can be splitted to
staged.  One of the future work is to explore that whether these
analyses can be implemented and optimized from
meta-programming. Souffl{\'e}, one of the state-of-the-art Datalog
engine that is commonly used in points-to analysis
\cite{Antoniadis:2017:PDS:3088515.3088522}, uses the idea of Futamura
projection for efficient implementation \cite{10.1007/978-3-319-41540-6_23}.
But it only specializes Datalog rules to the running engine, instead
of a full pipeline from program AST to generating constraints and
solving constraints.

Recent work shows that the numerical abstract domains can be
specialized w.r.t. the static program structure, although not in the
way of using staging. For example, decomposing polyhedrons
\cite{DBLP:conf/popl/SinghPV17, Singh:2017:PCD:3177123.3158143} to
smaller ones that depend on the variables involved in a statement can
significantly improve the running time and space of abstract
transformers.

@@

\paragraph{Control-flow Analysis} Our case studies and evaluation focus on
control-flow analysis for functional languages
\cite{Shivers:1991:SSC:115865.115884, Midtgaard:2012:CAF:2187671.2187672}. In
general, $k$-CFA (where $k > 0$) is considered impractical to use as a compiler
pass, due to its EXPTIME lower bound \cite{VanHorn:2008:DKC:1411204.1411243}.
However, several variants of 0CFA were developed and shown to be useful in
practice and program optimization \cite{Adams:2011:FTR:2048066.2048105,
Bergstrom:2014:PEH:2628136.2628153, ashley:practical, Reppy:2006:TCA:1159876.1159888}.

\paragraph{Two-level Semantics} The idea of reinterpreting the semantics as
abstract interpretation can be traced to Nielson's two-level semantics
\cite{NIELSON1989117}; using two-level semantics for code generation
was also explored by \citet{NIELSON198859}.
\citet{Sergey:2013:MAI:2491956.2491979}'s work of monadic abstract
interpreters is also closely related to the two-level semantics: the
use of a generic interface with monads and then reinterpret it by
different semantics is already two-level. Instead of focusing on
semantics, this paper shows how a staged analyzer can be built and
used to increase efficiency of static analysis, we augment the monadic
abstract interpreter from another dimension, using code generation to
produce efficient low-level code. Hence the presented work can be
considered as an two-dimension application of the two-level semantics.
